---
title: MPD publication details
subtitle: Draft
type: post
---

```{r var-yaml-code, eval=FALSE, include=FALSE}
#     theme: cerulean
#     csl: mee.csl

# bibliography: references.bib
  # html_document:
  #   fig_caption: yes
  #   highlight: pygments
  #   number_sections: no
  #   theme: cerulean
  # pdf_document:
  #   fig_caption: yes
  #   keep_tex: yes
  #   number_sections: yes
  # word_document: default
  
```

```{r enviroment, include=FALSE}
# libraries
  library(tidyverse)
  library(lubridate)
  library(jagsUI)
  library(knitr)
  library(gridExtra)
  library(extrafont)
  library(gdata)

# sorting fonts
  windowsFonts()
  loadfonts(device = "win")
  
# this just simply reduces long name to TIMES as in code below
  windowsFonts(Times=windowsFont("TT Times New Roman")) 
  
# devtools::install_github("cboettig/knitcitations@v1")
  library(knitcitations); cleanbib()
  cite_options(citation_format = "pandoc", check.entries=FALSE)
  library(bibtex)
```

## Publishing notes

Journal choice: [Journal of Applied Ecology](http://www.journalofappliedecology.org/view/.../authorGuideline.html)

- [Author Guidelines](http://mc.manuscriptcentral.com/jappl-besjournals)
  - A research **"article"** should not exceed **7000** words (the word count is inclusive of all parts of the manuscript, including the title page, abstract, references, table and figure legends, but excluding files uploaded as Supporting Information). See Manuscript specifications below.

- [Assistant editor's email](admin@journalofappliedecology.org)

- [Submission checklist](https://besjournals.onlinelibrary.wiley.com/hub/journal/13652664/about/author-guidelines#quickchecklistinitialsubmission)

##### Title page

##### Abstract

##### Keywords

##### Introduction

##### Materials and methods

##### Results

##### Discussion

##### Conclusions (optional)

##### Authors’ contributions

##### Acknowledgements (optional)

##### Data accessibility

##### References
